sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2023-09-12 doc/paper-jss/paper-jss.Rmd
\boldsymbol{} -> \bm{}

The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects $\boldsymbol{\beta}$ and offset $O_{\boldsymbol{s},t}$, IID (independent and identically distributed) random intercepts $\alpha_g$, spatial intercept random fields $\omega_{\boldsymbol{s}}$, spatiotemporal intercept random fields $\epsilon_{\boldsymbol{s},t}$, time-varying effects $\boldsymbol{\gamma_t}$, and spatially varying effects $\zeta_{\boldsymbol{s}}$ (Figure \ref{fig:diagram}). These components are discussed in more detail via our examples below.

$$
\begin{aligned}
\mu_{\boldsymbol{s},t} &=
f^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s},t} \boldsymbol{\beta} +
O_{\boldsymbol{s},t} +
\alpha_g +
\omega_{\boldsymbol{s}} +
\epsilon_{\boldsymbol{s},t} + 
\boldsymbol{X}^{\mathrm{tvc}}_{\boldsymbol{s},t} \boldsymbol{\gamma_t} +
\boldsymbol{X}^{\mathrm{svc}}_{\boldsymbol{s},t} \zeta_{\boldsymbol{s}} \right),
\end{aligned}
$$
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link function $f^{-1}$ (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.
See the model description vignette for a full description of the statistical model (`browseVignettes("sdmTMB")` or see the \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).

@@

$$
\begin{aligned}
y_{\boldsymbol{s}} &\sim \operatorname{Bernoulli} \left(\mu_{\boldsymbol{s}}\right),\\
\mu_{\boldsymbol{s}} &= \operatorname{logit}^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s}} \boldsymbol{\beta} +
\omega_{\boldsymbol{s}}
\right),
\end{aligned}
$$

where $\boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s}}$ contains main effect covariates (intercept, quadratic effects of depth), $\boldsymbol{\beta}$ represents estimated main effect coefficients, and $\omega_{\boldsymbol{s}}$ represents the estimated spatial field,

$$
\omega_{\boldsymbol{s}} \sim \mathrm{MVN}(\boldsymbol{0},\boldsymbol{\Sigma}_\omega)
$$

@@

$$
\begin{aligned}
\mathbb{E}[y_{\boldsymbol{s},t}] &= \mu_{\boldsymbol{s},t},\\
\mu_{\boldsymbol{s},t} &=
f^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s},t} \boldsymbol{\beta} +
O_{\boldsymbol{s},t} +
\boldsymbol{X}^{\mathrm{tvc}}_{\boldsymbol{s},t} \gamma_t +
\omega_{\boldsymbol{s}} +
\epsilon_{\boldsymbol{s},t} 
\right),
\end{aligned}
$$

and TODO

@@

$$
\begin{aligned}
\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon_{t}},  \:
\boldsymbol{\epsilon_{t}} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
\end{aligned}
$$

@@

$$
\begin{aligned}
\mathbb{E}[y_{\boldsymbol{s},t}] &= \mu_{\boldsymbol{s},t},\\
\mu_{\boldsymbol{s},t} &=
f^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s},t} \boldsymbol{\beta} +
\alpha_g +
\boldsymbol{X}^{\mathrm{svc}}_{\boldsymbol{s},t} \zeta_{\boldsymbol{s}} +
\omega_{\boldsymbol{s}} +
\epsilon_{\boldsymbol{s},t} \right),
\end{aligned}
$$

>>>>

The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects $\bm{\beta}$ and offset $O_{\bm{s},t}$, IID (independent and identically distributed) random intercepts $\alpha_g$, spatial intercept random fields $\omega_{\bm{s}}$, spatiotemporal intercept random fields $\epsilon_{\bm{s},t}$, time-varying effects $\bm{\gamma_t}$, and spatially varying effects $\zeta_{\bm{s}}$ (Figure \ref{fig:diagram}). These components are discussed in more detail via our examples below.

$$
\begin{aligned}
\mu_{\bm{s},t} &=
f^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
O_{\bm{s},t} +
\alpha_g +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} + 
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma_t} +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} \right),
\end{aligned}
$$
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link function $f^{-1}$ (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.
See the model description vignette for a full description of the statistical model (`browseVignettes("sdmTMB")` or see the \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).

@@

$$
\begin{aligned}
y_{\bm{s}} &\sim \operatorname{Bernoulli} \left(\mu_{\bm{s}}\right),\\
\mu_{\bm{s}} &= \operatorname{logit}^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s}} \bm{\beta} +
\omega_{\bm{s}}
\right),
\end{aligned}
$$

where $\bm{X}^{\mathrm{main}}_{\bm{s}}$ contains main effect covariates (intercept, quadratic effects of depth), $\bm{\beta}$ represents estimated main effect coefficients, and $\omega_{\bm{s}}$ represents the estimated spatial field,

$$
\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)
$$

@@

$$
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
f^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
O_{\bm{s},t} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \gamma_t +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} 
\right),
\end{aligned}
$$

where $\bm{\beta}$ are coefficients associated with the main effects, $O_{\bm{s},t}$ represents the offset (here, log area swept), $\gamma_t$ represents the time-varying coefficients, $\omega_{\bm{s}}$ is the spatial field (equivalent to a spatial intercept) and $\epsilon_{\bm{s},t}$ represents spatiotemporal fields.

The temporally varying intercepts $\gamma_t$ are modeled as a stationary AR(1) process,

@@

$$
\begin{aligned}
\bm{\delta}_{t=1} &\sim \operatorname{MVNormal} (\bm{0}, \bm{\Sigma}_{\epsilon}),\\
\bm{\delta}_{t>1} &= \rho \bm{\delta}_{t-1} + \sqrt{1 - \rho^2} \bm{\epsilon_{t}},  \:
\bm{\epsilon_{t}} \sim \operatorname{MVNormal} \left(\bm{0}, \bm{\Sigma}_{\epsilon} \right),
\end{aligned}
$$

@@

$$
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
f^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\alpha_g +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right),
\end{aligned}
$$
